---
title: "Mussels data analysis"
author: "Your collaborator"

output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_section: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(readr)
library(psych)
```

# Reading the data

Dear Colleagues, congratulations on your successful expedition! 
In response to your request, here is some initial analysis to help you undersand your findings.

In order to start calculations, please provide the full path to the directory where your datasets are stored (that is, files with .csv extension) to a specific function made for you to help you integrate all your datasets at once. 

```{r, echo=FALSE}
read_dataset <- function(path_to_dataset) {
#dir <- ("Data")
  dir <- (path_to_dataset)
  files <- list.files(file.path(dir), pattern="*.csv", full.names = T)
  data_list = lapply(files, read.csv, header = TRUE, fill = TRUE)
  df_raw_data <- do.call(rbind, data_list)
  rm(dir,files,data_list)
  return(df_raw_data)
}
```

```{r}
df_raw_data <- read_dataset("~/BI/R/project1/Data") # For example, "~/BI/R/project1/Data"
```
Now, let's look at the row data and learn its main parametres:
```{r}
head(df_raw_data)
nrow(df_raw_data) # how many objects are in the dataset?
ncol(df_raw_data) # how many characteristics do they have?
```
Look in details on the structure of the dataset:
```{r}
str(df_raw_data) 
```

# Modification of the dataset structure

To begin with, as R failes to determine automatically types for the first three variables,
these variables seem to have some discordance. Let's look at them and do transformations where nesessary:
```{r}
levels(factor(df_raw_data$Rings)) # is there smth unusual with the levels for variable "Rings"?
levels(factor(df_raw_data$Sex..1...male..2...female..3...uvenil.)) # is there smth unusual with the second variable?
levels(factor(df_raw_data$Length)) # is there smth unusual with the levels for variable "Length"?
```
Let's fix this discordance:
```{r}
df_raw_data$Rings[df_raw_data$Rings == "nine"] <- 9 # translate the value "nine" into numeric data
names(df_raw_data)[names(df_raw_data) == 'Sex..1...male..2...female..3...uvenil.'] <- 'Sex' # change the name of the third variable and reset in values
df_raw_data$Sex[df_raw_data$Sex == "male"] <- "Male"
df_raw_data$Sex[df_raw_data$Sex == "1"] <- "Male"
df_raw_data$Sex[df_raw_data$Sex == "one"] <- "Male"
df_raw_data$Sex[df_raw_data$Sex == "2"] <- "Female"
df_raw_data$Sex[df_raw_data$Sex == "3"] <- "Juvenile"
df_raw_data$Sex[df_raw_data$Sex == "three"] <- "Juvenile"
df_raw_data$Length[df_raw_data$Length == "No data! I forgot to mesure it!("] <- NA

# levels(factor(df_raw_data$Rings)) # Use this commands to check the improvements if nessesary 
# levels(factor(df_raw_data$Sex))
# levels(factor(df_raw_data$Length))
```
Now, reset valid types for the first three variables and look for renewed dataset:
```{r}
df_raw_data$Rings <- as.numeric(df_raw_data$Rings) 
df_raw_data$Sex <- as.factor(df_raw_data$Sex)
df_raw_data$Length <- as.numeric(df_raw_data$Length)
str(df_raw_data)
```

# Exploratory analysis

Great! Let's start with EDA and analyse distributions of all numeric variables for the three sex levels of mussels:

```{r, echo=FALSE, warning=FALSE}
var1 <- ggplot(df_raw_data, aes(Sex, Rings, fill = Sex)) + 
  geom_violin(alpha=0.3) +
  geom_boxplot(width=0.2, alpha=0.8) +
  theme_bw() + 
  xlab(label = "Sex") + 
  ylab(label = "Rings") + 
  theme(legend.position = "None")

var3 <- ggplot(df_raw_data, aes(Sex, Length, fill = Sex)) + 
  geom_violin(alpha=0.3) +
  geom_boxplot(width=0.2, alpha=0.8) +
  theme_bw() + 
  xlab(label = "Sex") + 
  ylab(label = "Length") + 
  theme(legend.position = "None")

var4 <- ggplot(df_raw_data, aes(Sex, Diameter, fill = Sex)) + 
  geom_violin(alpha=0.3) +
  geom_boxplot(width=0.2, alpha=0.8) +
  theme_bw() + 
  xlab(label = "Sex") + 
  ylab(label = "Diameter") + 
  theme(legend.position = "None")

var5 <- ggplot(df_raw_data, aes(Sex, Height, fill = Sex)) + 
  geom_violin(alpha=0.3) +
  geom_boxplot(width=0.2, alpha=0.8) +
  theme_bw() + 
  xlab(label = "Sex") + 
  ylab(label = "Height") + 
  theme(legend.position = "None")

var6 <- ggplot(df_raw_data, aes(Sex, Whole_weight, fill = Sex)) + 
  geom_violin(alpha=0.3) +
  geom_boxplot(width=0.2, alpha=0.8) +
  theme_bw() + 
  xlab(label = "Sex") + 
  ylab(label = "Whole_weight") + 
  theme(legend.position = "None")

var7 <- ggplot(df_raw_data, aes(Sex, Shucked_weight, fill = Sex)) + 
  geom_violin(alpha=0.3) +
  geom_boxplot(width=0.2, alpha=0.8) +
  theme_bw() + 
  xlab(label = "Sex") + 
  ylab(label = "Shucked_weight") + 
  theme(legend.position = "None")

var8 <- ggplot(df_raw_data, aes(Sex, Viscera_weight, fill = Sex)) + 
  geom_violin(alpha=0.3) +
  geom_boxplot(width=0.2, alpha=0.8) +
  theme_bw() + 
  xlab(label = "Sex") + 
  ylab(label = "Viscera_weight") + 
  theme(legend.position = "None")

var9 <- ggplot(df_raw_data, aes(Sex, Shell_weight, fill = Sex)) + 
  geom_violin(alpha=0.3) +
  geom_boxplot(width=0.2, alpha=0.8) +
  theme_bw() + 
  xlab(label = "Sex") + 
  ylab(label = "Shell_weight") + 
  theme(legend.position = "None")

ggarrange(var1, var3, var4, var5, var6, var7, var8, var9,
                    ncol = 3, nrow = 3)
```

As one can see, "Sex" variable contain some missing value. How many such values?
```{r}
sum(is.na(df_raw_data$Sex))
```
There is only one such an observation. As can be seen from the figure above, this value can not be set manually. All that we could say about this value is just that it is likely to be from an adult individual. But it is impossible to discriminate whether it is a male or a female. So, we remove this observation from the dataset:
```{r}
df_raw_data <- dplyr::filter(df_raw_data, !is.na(Sex)) # remove NA from Sex
```
Next, check other variables for missing values:
```{r}
sum(is.na(df_raw_data$Rings))
sum(is.na(df_raw_data$Length))
sum(is.na(df_raw_data$Diameter))
sum(is.na(df_raw_data$Height))
sum(is.na(df_raw_data$Whole_weight))
sum(is.na(df_raw_data$Shucked_weight))
sum(is.na(df_raw_data$Viscera_weight))
sum(is.na(df_raw_data$Shell_weight))
```
Now, the dataset appears to have a relatively small number of missed values or severe outliers. In this situation, the best way to manage them is to replace for the median considering "Sex" grouping, as the median estimate better suits a distribution stably than the mean one and values are vary depending on "Sex" level.
```{r}
df_data <- df_raw_data %>%
  group_by(Sex) %>% 
  mutate(Length = ifelse(is.na(Length), median(Length, na.rm = T), Length)) %>% 
  mutate(Height = ifelse((Height > 0.3) | (is.na(Height)), median(Height, na.rm = T), Height)) %>% 
  mutate(Diameter = ifelse(is.na(Diameter), median(Diameter, na.rm = T), Diameter)) %>%
  mutate(Whole_weight = ifelse(is.na(Whole_weight), median(Whole_weight, na.rm = T), Whole_weight)) %>%
  mutate(Shucked_weight = ifelse(is.na(Shucked_weight), median(Shucked_weight, na.rm = T), Shucked_weight)) %>%
  mutate(Viscera_weight = ifelse(is.na(Viscera_weight), median(Viscera_weight, na.rm = T), Viscera_weight)) %>%
  mutate(Shell_weight = ifelse(is.na(Shell_weight), median(Shell_weight, na.rm = T), Shell_weight)) 
sum(is.na(df_data$Length)) # check transformation for "Length" variable
```
Finally, let's look at the interrelation between all numeric variables of the fixed dataset:
```{r}
df <- df_data[, -c(2)]
plot(df)
```
As can be seen, in general, almost all pairs of variables have some degree of connection, espesially those ones that consider close parameters (size or weight). Here we can also see, that our transformations were successful, as the variable "Height" does not contain obvious outliers anymore.

Here is also main statistics of all variables, depending on mussels sex:
```{r}
describeBy(df_data[,c(-2)], group = df_data$Sex)
```
# Specific questions

## Mean and sd of mussels "Length" depending on "Sex"

```{r}
df_data %>% 
  group_by(Sex) %>%  
  summarise(Mean = mean(Length),
            SD = sd(Length))
```

## Percentage of molluscs that has the "Height" variable less than 0.165

```{r}
num_of_subset <- length(df_data$Height[df_data$Height <= 0.165])
num_total <- length(df_data$Height)
persentage <- 100 * num_of_subset / num_total
round(persentage, 2)
```

## "Length" value, which is greater than 92% of all observations

```{r}
quantile(df_data$Length, probs = c(0.92))
```

Let's look at this:

```{r}
ggplot(df_data, aes(Length)) + 
  geom_histogram(alpha=0.6) +
  geom_vline(xintercept=0.67, linetype="dashed", 
             color = "red", size=1)+
  theme_bw()
```

## Generating new variable "Lenght_z_scores" by "Length" values standardization

```{r}
df_data <- df_data %>% 
  mutate(Length_z_scores = (Length - mean(Length)) / sd(Length))
head(df_data)
```

## Comparison of the diameters of the mussels with the number of rings 5 and 15

First, subset the dataset to get observations relevant for the comparison, convert "Rings" variable to a factor and look at the data:
```{r}
df7 <- df_data %>% 
  filter(Rings == c(5, 15))
df7$Rings <- as.factor(df7$Rings)
str(df7)
```

Next, let's visualize a difference for the compared groups. Here are some additional ready for a publication plots made for you. Notably, nowadays, despite being a bit nonstandard, the second plot is considered to be the most informative, as it reflects both a degree of confidence (distanse of CIs) and distrubutions of values of the compared groups:

```{r}
ggplot(df7, aes(Rings, Diameter, fill=Rings)) +
  geom_violin(alpha=0.3) +
  geom_boxplot(width=0.2, alpha=0.8) +
  #stat_summary(fun.data=mean_se, geom="errorbar", color="black", width=0.3) +
  theme_bw() + 
  xlab(label = "Rings number") + 
  ylab(label = "Diameter") + 
  theme(legend.position = "None")

ggplot(df7, aes(Rings, Diameter, fill=Rings)) +
  geom_violin(alpha=0.3) +
  stat_summary(fun = mean, geom = "point") + 
  stat_summary(fun.data = mean_se, geom = "errorbar", width=0.3) +
  theme_bw() + 
  xlab(label = "Rings number") + 
  ylab(label = "Diameter") + 
  theme(legend.position = "None")
```

It is seen, that it has to be a huge difference between groups. Let's chech it more strictly by testing via t-test. Firstly, check for the conditions of applicability such a parametric criteria, and then run the test:

```{r}
ggplot(df7, aes(Diameter, fill = Rings)) + # visualization of distributions
  geom_density(alpha=0.7) + 
  theme_bw()

shapiro.test(df7$Diameter[df7$Rings == "5"]) # whether we can say the values are distributed unnormally
shapiro.test(df7$Diameter[df7$Rings == "15"])

bartlett.test(Diameter ~ Rings, df7) # is there a difference between variation between the groups

t.test(Diameter ~ Rings, df7) 
```

## Interrelation between "Diametr" and "Whole_weight" variables

First, make a scaterplots for the variables. Secong plot, on the left, has been built via sqrt() transformation of the dependant variable, as the connection between variables is not linear:

```{r, }
plot1 <- ggplot(df_data, aes(Diameter, Whole_weight, fill = "white")) +
  geom_point(alpha=0.3) + 
  geom_smooth() + 
  theme_bw() + 
  xlab(label = "Diameter") + 
  ylab(label = "Whole weight") +
  theme(legend.position = "None")

plot2 <- ggplot(df_data, aes(Diameter, sqrt(Whole_weight), fill = "white")) +
  geom_point(alpha=0.3) + 
  geom_smooth(method = "lm") + 
  theme_bw() + 
  xlab(label = "Diameter") + 
  ylab(label = "Whole weight") +
  theme(legend.position = "None")

ggarrange(plot1, plot2, ncol = 2, nrow = 1)
```

To check statistically apply Pearson's correlation as follows:

```{r}
cor.test(sqrt(df_data$Whole_weight), df_data$Diameter, method = 'pearson')

```

Then we can create a linear model describing the interconnection of variables. As can be seen, tha variables a positively and statistically significantly connected:
```{r}
model <- lm(sqrt(Whole_weight) ~ Diameter, df_data)
summary(model)
```

In addition, one can look in more details at the analysed interconnection, as the model for juvenile individuals might be slightly different from the adult ones, which are almost simular:

```{r}
ggplot(df_data, aes(Diameter, sqrt(Whole_weight), color = Sex)) +
  geom_point(alpha=0.3) + 
  geom_smooth(method = "lm") + 
  theme_bw() + 
  xlab(label = "Diameter") + 
  ylab(label = "Whole weight") +
  scale_color_manual(values=c("#5a5255", "#c3cb71", "#559e83"))+
  theme(legend.position = "bottom")
```














